A Linear Solution of the Linearity Challenge

### 2025-12-21 paper/introduction.tex
intro

One of the challenges in the benchmark, henceforth called \emph{linearity
challenge}, concerns the formalization of a minimal \emph{session calculus}.
%
Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
are established formalisms for the static analysis of distributed programs based
on peer-to-peer communication channels. Every session type system revolves
around three key ideas: (1) session endpoints are \emph{linear resources} that
cannot be discarded or duplicated without compromising some safety and liveness
properties of a program; (2) the type of a session endpoint is \emph{updated}
after each use to reflect the state of the interaction it describes; (3) peer
session endpoints are meant to be used in complementary ways so as to guarantee
the absence of communication errors and, to some extent, progress of the
interaction.
%
The linearity challenge is motivated by the observation that the proper handling
of linear resources in a formalization often requires a large number of complex
auxiliary definitions and technical results that divert from the main problem
under consideration. One of the alleged culprits of such unwanted complexity is
\emph{context splitting}, namely the operation that partitions a typing context
in such a way that the linear resources described therein end up in only one of
the partitions. This observation has led to the development of various
techniques for dealing with typing contexts without splitting them, including
leftover typing~\cite{ZalakainDardha21}, the use of linearity
predicates~\cite{SanoKavanaghPientka23} and contexts with
tags~\cite{ZackonEtAl25}.

This paper presents a solution to the linearity challenge.
%
The calculus that we formalize, which we call \Calculus, and the one in the
challenge description differ for two key aspects.
%
First of all, the challenge only considers a minimal calculus of first-order
sessions, whereas \LCP is a fully-featured calculus that includes support for
higher-order, polymorphic sessions and both linear and shared channels. Also, as
suggested by its name, \LCP is closely related to \CP~\citep{Wadler14} and its
type system is based on classical linear logic.
%
The second difference is that the calculus in the challenge is equipped with
\emph{native} sessions, whereas \LCP is a \emph{linear} calculus where each
channel is meant to be used for a \emph{single} communication. Structured
sessions involving the exchange of several messages can be modeled thanks to the
use of \emph{explicit continuation channels}. This encoding of (binary) sessions
using linear channels is well known in the
literature~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP has the same expressive power of \CP, but is a lower-level
calculus in which sessions can be encoded instead of being featured natively.

The formalization we present is the most complete (in terms of features
supported by the calculus) and, at the same time, the simplest one (in terms of
code size). It is also one of only two formalizations that prove the deadlock
freedom property, and it does so with an order of magnitude less code than the
existing formalization~\cite{JacobsBalzerKrebbers22}.

>>>>

One of the challenges in the benchmark, henceforth called \emph{linearity
challenge}, concerns the formalization of a \emph{session calculus}.
%
Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
are established formalisms for the static analysis of distributed programs based
on peer-to-peer communication channels. Every session type system revolves
around three key ideas: (1) session endpoints are \emph{linear resources} that
cannot be discarded or duplicated without compromising some safety and liveness
properties of a program; (2) the type of a session endpoint is \emph{updated}
after each use to reflect the state of the interaction it describes; (3) peer
session endpoints are meant to be used in complementary ways so as to guarantee
the absence of communication errors and, to some extent, progress of the
interaction.
%
The linearity challenge is motivated by the observation that the proper handling
of linear resources in a formalization often requires a large number of complex
auxiliary definitions and technical results that divert from the main problem
under consideration. One of the alleged culprits of such unwanted complexity is
\emph{context splitting}, namely the operation that partitions a typing context
in such a way that the linear resources described therein end up in only one of
the partitions. This observation has led to the development of various
alternatives to context splitting including leftover
typing~\cite{ZalakainDardha21}, the use of linearity
predicates~\cite{SanoKavanaghPientka23} and tagged contexts~\cite{ZackonEtAl25}.

This paper presents a solution to the linearity challenge.
%
The calculus that we formalize, dubbed \Calculus, and the one in the challenge
description differ for two key aspects.
%
First of all, the challenge only considers a minimal calculus of first-order
sessions, whereas \LCP is a fully-featured calculus that includes support for
higher-order, polymorphic, linear and shared channels. Also, \LCP is closely
related to \CP~\citep{Wadler14} and its type system coincides with the proof
system of classical linear logic.
%
The second difference is that the calculus in the challenge is equipped with
\emph{native} sessions, whereas \LCP features shared channels, which are
unrestricted, and linear channels, which are meant to be used for a
\emph{single} communication. Structured sessions involving the exchange of
several messages can be modeled using \emph{explicit continuations} according to
the well-known encoding of (binary) sessions into linear
channels~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP has the same expressive power of \CP, but is a lower-level
calculus in which sessions are encodable instead of being featured natively.

Compared to the other analogous formalizations of session/linear
calculi~\cite{GotoEtAl16,Thiemann19,CicconePadovani20,CastroFerreiraYoshida20,RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,SanoKavanaghPientka23,ZackonEtAl25},
the one we present here is both the most complete (in terms of features
supported by the calculus) and the most streamlined (in terms of code size). It
is also one of only two formalizations that prove the deadlock freedom property
for a session calculus, and it does so with substantially less code than the
existing formalization~\cite{JacobsBalzerKrebbers22}.

We acknowledge that \LCP is not

The differences between \LCP and the calculus are functional to the simplicity
of its formalization.
